DRAFT: Harmonized and trait aggregation paper

### 2020-12-08 Paper/Draft_harmonization_paper.tex
08.12: Discussion updates

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from various trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising trait information from multiple trait databases showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to amalgamate these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Thus, the field of freshwater ecology needs terminological standards to facilitate data synthesis. Harmonised definitions and concepts of traits have been for example developed for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is not only to provide standardized trait definitions but also to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater science to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could then be linked through standardised terminologies or ontologies, as suggested by Baird et al. almost a decade ago \cite{baird_toward_2011}. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately improve interpretation of derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

%%% No section on codings?
% Besides semantic heterogeneity between traits and grouping features, codings used to describe traits differ between trait databases. % We showed that fuzzy coded and binary coded could be used together but a lot of assumptions have been made here?
%%% Section to different codings -> what is an appropriate question?
% Maybe this section is not necessary and just one or two sentences to various codings is sufficient
% The European and Australian trait databases use several codings for their traits (fuzzy, binary and continuous) while the North American databases only use binary coding and the New Zealand database only fuzzy coding. % due to a lack of species trait information.
% Differences in fuzzy coding:
% freshecol: 
% It is based on the average distributions, occurrences or behaviours of a taxon within the environmental gradient or organismic flexibility under consideration. The ten point assignment system was first introduced by Zelinka and Marvan (1961) for the classification of saprobic valences and was extended by Moog (1995) to other parameters, such as stream zonation preferences and feeding types. Within freshwaterecology.info the ten point assignment system was used for coding the majority of those parameters, which comprise several categories.
% Example for the parameter "stream zonation preference": If 70% of a species’ records are observed in spring brooks and 30% in the upper trout region, 7 out of 10 points will be allocated to the category "hypocrenal" (spring brook) and 3 points to "epirhithral" (upper trout region) to describe the expected occurrence of this species within the longitudinal zonation of a river. For all other categories "0" will be used.
% Tachet: 0 - 5: Thereby, the relevant features of a taxon were described by the relative distribution of the information among the categories of each trait. This procedure was a pertinent and simple way to "translate" the available knowledge about species traits of organisms into numerical values. It also avoided the obligate assignment of a taxon to a single category, which can lead to inaccurate characterisation of biological or ecological trait profiles
% Can be converted so that they can be used together 
% As already stated fuzzy coding is used to express uncertainty in traits, for variability in traits within taxononomic groups above species-level or for variation in traits during the development of an organism. So it's not really clear for fuzzy coded taxa on family-level what uncertainty is expressed
% Fuzzy coding captures uncertainty. Nice way but has also drawbacks -> correlated variables, needs specific statistic methods, cannot be used in bayesian priors etc.
% Binary coding is, transformed to 0 and 1, and thus converted just a "variant" of fuzzy coding?
% Problems fuzzy coding: Which uncertainty do they cover?
% Stay focused! Schneide nicht zu viele Themen an - Du musst nicht jedes einzelne Thema diskutieren. Generell kann die Diskussion so 3-4 Abschnitte haben, die jeweils die wichtigsten Resultate oder Aspekte diskutieren. Eine generelle Diskussion der Fuzzy Coding und intra-species variation wäre aus meiner Sicht eher ein eigener, wenn auch nicht allzulanger Abschnit
% \textit{\color{blue} Ideas for concluding this section:
% \begin{itemize}
%   \item The uncertainty of trait information should be taken into account when aggregating traits and appropriate measures to capture this uncertainty should be established.
%   \item Are other measures of uncertainty thinkable, e.g. based on SD or Variance 
% \end{itemize}
% }

%%%%%%% Section on taxonomic levels and data coverage
%? What's the problem with the state of data coverage
Our analysis aimed to use most of the available trait information for different regions to establish harmonised grouping feature datasets. Although many taxa are coverd in the trait databases, the availability of trait information varies strongly across grouping features and taxonomic resolution differs between databases.
% freshwaterecolgy.info on species-level, Australia mainy taxa on genus and family-level
% Pointing to the problem of how accurate information on levels with lower taxonomic precision is 
% For some groups, species and genus-level identification important since information varies considerably in these groups; see Serra Paper!
% Erwähnen? Obtaining trait information on species-level is challenging
% How would this change our results?

@@

% R: Man könnte das ggf. auch an den Anfang ziehen. Also einleitend: Warum ist die Frage wichtig, was haben wir überhaupt gemacht - und dann eben erst mal sagen, dass die Analysen zwischen aggregated und assigned traits jetzt nicht die ultimative Wahrheit sind. Gleichwohl unsere Ergebnisse anzeigen können, welche Methoden überhaupt und unter welchen Umständen zu unterschiedlichen Ergebnissen führen, so dass die Leser dort entsprechend vorsichtig sein müssten

As outlined in the introduction, various trait aggregation methods are used by researchers when trait data are on more precise taxonomic levels than observational data. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision which is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction \cite{monaghan_improving_2013}. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) aggregation of traits using \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{mean}} (median aggregation methods) compared to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_mean} (mean aggregation methods) is often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). However, when there were differences between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American trait dataset. The mean absolute differences for the median aggregations was twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits for North America. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. Thus, using the median led for the North American dataset in many cases either to agreement with the assigned traits or the difference between median aggregated traits and assigned traits was often 1. 

@@

The number of differing cases and the mean absolute differences to the assigned traits were similar within the mean aggregation and within the median aggregation methods, which confirms our second finding, the small influence of weighting on the aggregation methods. Furthermore, the distributions of absolute trait affinity differences to assigned traits obtained were relatively similar for the mean aggregation methods and the median aggregation methods in both tested datasets (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset to the assigned traits at family-level, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}).

@@

% R: Allgemein in Trait data sets - ja geht in die Richtung, wobei du ja hier nicht noch analysen für alle möglichen Tiere und traits machen kannst, aber zumindest aufzeigen unter welchen Umständen es bedeutsam ist - und was dann gemacht werden sollte. Der Leser will ja aus der Diskussion etwas mitnehmen, ggf. auch was er tun soll
\textit{\color{blue} Maybe include: How often does such a taxonomic data occur in the trait datasets?}
% Das liest sich wie einer erweiterte Resultate Sectioin - das ist nicht Sinn der Diskussion. Du sollst da nicht neue Ergebnisse vorstellen sondern dann eher auf eine Abbildung verweisen und in der Regel in 2-3 Sätzen die Ergebnissen in einfachen Sätzen nennen und diskutieren d.h. in Kontext anderer Studien stellen - oder wenn es diese nicht gibt, was auch gesagt werden müsste, dann sagen, was daraus folgt.

% R: Misses a starting sentence, Why did we conduct this analysis?
In the re-analysis where we investigated how using harmonised and aggregated traits might change the results of trait-environment relationships we found similar results compared the original study. Traits that responded only in the original analysis but were non-response in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see methods), namely feeding mode shredder, respiration gills, and life cycle duration traits. For example, during harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original dataset have a higher mean and standard deviation than in the aggregated and harmonized datasets (Table S\ref{tab:SI_resp_traits_summary_stats}). Hence, the signal in the original data for the trait shredder has presumably been weakened by the harmonization, which might be the reason that shredders showed a less strong reaction to the salinity gradient.

@@

% What do we really know because someone has watched the animals and published their behaviour; what is extrapolated knowledge? Where does the classification come from?
% quality evaluation of traits als Thema..

>>>>

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Although, others have noted the lack of standardized trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. To resolve definition discrepancies and facilitate data synthesis terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardized trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater science to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could then be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately improve interpretation of derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

% What's the problem with the state of data coverage
Our analysis aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets. Although many taxa are covered in the trait databases from which we extracted our trait datasets, the availability of trait information varies strongly across grouping features and taxonomic resolution differs between databases. While trait information from Europe and New Zealand was mainly on species-level, considerable portions of trait data from North America and Australia were on genus or family-level (Table \ref{tab:tax_coverage}). Assigned traits on family-level might not reflect the real trait diversity within a taxon, for example for Chironomidae \cite{serra_synthesising_2016}. However, species identification can be complicated and time-consuming \cite{marshall_taxonomic_2006, resh_which_2008}, and therefore trait information on family-level has been widley used e.g. in bioassessment \cite{beketov_spear_2009}.
% How would obtaining information on species level change our results?

@@

In the case when trait values are only available at species-level and observational data are on less precise taxonomic levels various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision which is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction \cite{monaghan_improving_2013}. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) aggregation of traits using \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{mean}} (median aggregation methods) compared to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_mean} (mean aggregation methods) is often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). However, when there were differences between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American trait dataset. The mean absolute differences for the median aggregations was twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits for North America. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. Thus, using the median led for the North American dataset in many cases either to agreement with the assigned traits or the difference between median aggregated traits and assigned traits was often 1. 

@@

The number of differing cases and the mean absolute differences to the assigned traits were similar within the mean aggregation and within the median aggregation methods, which confirms our second finding, the small influence of weighting on the aggregation methods. Furthermore, the distributions of absolute trait affinity differences to assigned traits obtained were relatively similar for the mean aggregation methods and the median aggregation methods in both tested datasets respectively (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset to the assigned traits at family-level, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}).

@@

% R: Misses a starting sentence, Why did we conduct this analysis?
% Mention that comparison to other studies not possible since they don't exist?
In the re-analysis where we investigated how using harmonised and aggregated traits might change the results of trait-environment relationships we found only slightly different results compared to the original study. Some of the traits that responded in the original analysis did not respond in the re-analysis. Those traits that responded only in the original analysis but were non-response in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see methods), namely feeding mode shredder, respiration gills, and life cycle duration traits. A reason for their non-responsiveness was likely the harmonisation procedure. For example, during harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original dataset have a higher mean and standard deviation than in the aggregated and harmonized datasets (Table S\ref{tab:SI_resp_traits_summary_stats}). Hence, the signal in the original data for the trait shredder has presumably been weakened by the harmonisation, which might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. The fact that the family-level aggregated traits showed similar results compared to the original analysis is in line with the findings of others that family-level traits can be sensitive enough to detect environmental impacts \cite{beketov_spear_2009}, although species-level traits are expected to result in more precise results when investigating trait-environment relationships. % Cite Schmidt Kloiber? Cite another paper that uses family-level trait data

@@

\section*{Conclusion}

% What do we really know because someone has watched the animals and published their behaviour; what is extrapolated knowledge? 
% Quality evaluation of traits als Thema..
% What do we learned: 
Comprehensive invertebrate trait databases have been developed over the last years in freshwater ecology. We showed that trait definitions are distinct across these databases. By providing an overview of definition discrepancies we hope to provide a starting point for the development of standardized trait definitions through which trait databases can be linked. Standardized terminolgy will facilitate comparable trait analysis across large scales. 
% - Variation in taxonomic coverage 
If trait aggregation is necessary we showed that using the median is often in conclusion with expert assignments. 
% Different weighting approaches did
Which method is objectively the a better measure cannot be answered, but traits with high variance should be treated carefully and taxa where a few genera have a disproportional high number of species trait information. If researchers need to aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate.

% Efforts in creating trait ontologies etc etc
% Technology improvement, e.g. Deep learning for species identification
